---
title: "LLM prompting with `ellmer` in R"
author: Hauke Licht
institute: University of Innsbruck
date: 2025-11-06
toc: true
toc-depth: 2
toc-location: right
toc-expand: true
execute:
  echo: true
  eval: true
  message: false
  warning: false
  cached: true
  fig-height: 2.8
  fig-width: 4.5
code-annotations: select
format:
  html:
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
---

```{r setup}
#| echo: false
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(scipen = 1, digits = 4) #set to two decimal 
```

# Introduction

## Ingredients

to prompt LLMs in R using the `ellmer` package, we need

1. an R environment with the `ellmer` package installed.
2. a "backend" LLM service that `ellmer` can connect to.

In this notebook, we will rely on Hugging Face's (HF) [_Inference Providers_](https://huggingface.co/docs/inference-providers/en/index) API.
But you can also use other backends supported by `ellmer`, such as OpenAI, Cohere, or local models via `ollama` (see [here](https://ellmer.tidyverse.org/reference/index.html#chatbots))

# Setup

## Package import

```{r load-packages}
library(ellmer)  # TODO: run renv::install("ellmer@0.4.0") if you get an error here
library(usethis) # TODO: run renv::install("usethis") if you get an error here
```

## Setting your hugging face API key

We will use the Hugging Face Inference Providers API as our LLM backend.
This means that the LLMs we prompt are running in the cloud.
We therefore use an [API](https://en.wikipedia.org/wiki/API) to send our prompts to the model and the API also sends back the requests to us.

To access this API, you need to insert the API key to the `.Renviron` file in your current _RStudio_ R project:

1. Open your R project in RStudio by running `usethis::edit_r_environ(scope = "project")`
2. In the file that is opened in RStudio, insert in a new line the text `HUGGINGFACE_API_KEY=`
3. Paste the API key I have posted in the **_OpenOLAT Forum_** your after the `=` sign (without any white spaces)
4. Save the file and restart R (Session -> Restart R in RStudio)
5. Verify that this succeeded by running the line
   ```r
   !is.na(Sys.getenv("HUGGINGFACE_API_KEY", unset = NA))
   ```
   If this returns `FALSE`, something went wrong. 
   Please ask me for help in this case.

::: {.callout-note title="cloud usage"}

Using models hosted by Hugging Face and its inference providers in the cloud means that

1. you need a _stable internet connection_ to use them 
2. you need to _pay_ for usage through an API account (see [here](https://huggingface.co/docs/inference-providers/en/pricing) and [here](https://huggingface.co/inference/models))

:::

::: {.callout-warning title="Costs"}

Generating responses with an LLM with the API token costs money. 
This money is charged on my account. 
I will cover the costs arising for your LLM usage during this course (up to $4 per person).

So please **be careful when using LLM in R with my API token**.
In particular:

1. Test your LLM text classification code always only with a handful of texts.
2. Avoid just clicking on "Run all" all the time and execute relevant lines selectively in your R scripts.
3. If running a code line or block that triggers an LLM API requests (e.g., `model$chat()`) results in an error, first make sure that you understand the source of the error instead of trying to run it again and again.

Should I see that someone uses the API token excessively, I'll ultimately need to revoke their access to it.


:::

::: {.callout-note title="Future use"}

In the future, you'll need to setup your own Hugging Face Pro account and pay for LLM usage  yourself (see [here](https://huggingface.co/pro) and [here](https://huggingface.co/settings/tokens))

:::

## Setting up the LLM backend

We first need to specify the model path and name of the LLM we want to use:

We'll use `Qwen3-235B-A22B-Instruct-2507` by `Qwen` from privider `novita`:

```{r model_name}
model_id <- "Qwen/Qwen3-Next-80B-A3B-Instruct:together"
```

::: {.callout-note title="Model choice"}

You find a list of available LLMs [here](https://huggingface.co/models?pipeline_tag=text-generation&sort=downloads&search=instruct).

Once you have found a suitable model, check its availability [here](https://huggingface.co/inference/models)

:::

Now we can setup the backend with `ellmer`'s `chat_huggingface()` function:

```{r}
# create a hugging face backend
model <- chat_huggingface(
  model = model_id,
  params = params(
    temperature = 0.0, 
    seed = 42
  )
)
class(model)
```

::: {.callout-note title="`ellmer` Chat objects"}

The `model` object created above is a `Chat` object.

To see the available methods defined for the `Chat` class, run: `help("Chat", package = "ellmer")`

:::

## Basic prompting

Now we can use the `model` object to prompt the LLM.
This is done by calling the backend's `chat()` method with a prompt string:

```{r basic-prompting}
model$chat("What is the capital of Austria? Respond in one sentence.", echo = "none")
```

As you see, this returns the model's response as a string.

# Text classification

::: {.callout-warning title="Always keep the convo tidied up"}

You need to always reset the conversation history of the LLM starting a new prompt session.
Otherwise, earlier "turns" between user input and model response will influence the models classification decisions

In ellmer, this can be done by running `model$set_turns(list())` before a new call of `model$chat()`:

```{r}
model$set_turns(list())
```

:::

To classify a text, we can instruct the LLM to do so in the prompt.
At a minimum, we

1. describe the classification task 
2. specify the possible label categories
3. instruct the model to only respond with the choosen category
4. and finally provide the text.


```{r}
instruction <- '
Classify the following text into positive, negative, or neutral sentiment. 
Only respond with one of these three labels.

Text:'
# remove line breaks at start and end 
instruction <- trimws(instruction)

# add to be classified text
input <- paste(instruction, '"""I love programming in R!"""')

model$chat(input)
```

::: {.callout-note title="Reseting the conversation history"}


`model$set_turns(list())` resets the conversation history to an empty list, meaning that no previous messages are considered by the LLM when responding to the new prompt.

:::

## Using the system message

As you know from my introduction, in LLM prompting, inputs and model responses are organized in a so-called **_conversation history_**

- each conversation history is composed a list of **_messages_**
- each message has two components
    1. a **_content_** (the text) _and_
    2. a role
- the relevant **_roles_** are:
    - _system_
    - _user_
    - _assistant_

This means that each conversation history is a list of messages, where each message has a content and a role.

- The **system message** is used to set the behavior of the LLM for the entire conversation. It is _optional_ and if specified, its only defined once at the beginning of a conversation history.
- The **user messages** are the actual prompts that you send to the LLM. In text classification applications, its the to-be-classified text, e.g., `"I love programming in R!"`.
- The **assistant messages** contains the responses of the LLM to the user messages. In text classification applications, its the predicted label, e.g., `"positive"`.

In the example above, we have specified the classification instruction and to-be-classified text all in one user message.

But we can also separate these components:


```{r}
#| echo: false
model$set_turns(list())
```


```{r}
instructions <- "
You will be provided with a text.

Your task is to classify the text's sentiment using the categories positive, negative, or neutral.

Only respond with one of the allowed labels: positive, negative, neutral
"

# set the system message of the model
model$set_system_prompt(instructions)

# specify the text to be classified as user message
text <- "I love programming in R!"

# send the text to the model
model$chat(text)
```

## Structured responses

To ensure that the LLM always only chooses from the allowed categories, we can specify a **response format**.
Response formats can be complex (see [here](https://ellmer.tidyverse.org/articles/structured-data.html)) but in classification, we only need to define the list of allowed label categories:

```{r}
response_format <- type_enum(
  values = c("positive", "neutral", "negative"), 
  description = "Sentiment classification of the input text"
)
```

We can pass this information when prompting the LLM to generate a classification response by using the `chat_structured()` method (instead of `chat()`) and passing the response format to the `type` argument:

```{r}
model$set_turns(list())
```

```{r}
# NOTE: we re-use the instructions defined above
model$set_system_prompt(instructions)

text <- "I love programming in R!"

# generate a structured response 
model$chat_structured(text, type = response_format)
```

You can see now that this time, we get a character value returned.

## Classifying multiple texts

To classify multiple texts at once using the same classification instructions, we can use `ellmer`'s `parallel_chat()` function.

This function accepts a `Chat` object (here, `model`) and a _list_ of input texts to be classified.
If we want structured responses, we use the `parallel_chat_structured()` function instead and also need to pass the response format to the `type` argument:


```{r}
model$set_turns(list())
```

```{r}
# NOTE: we re-use the instructions defined above
model$set_system_prompt(instructions)

texts <- c(
  "I love programming in R!",
  "Learning text analysis with Hauke is fun!",
  "But I hate bugs in my code.",
  "Oh well, debugging is okay, I guess."
)

parallel_chat_structured(
  model,
  prompts = as.list(texts), # NOTE: important to convert to list
  type = response_format # NOTE: we re-use the response format defined above
)
```

::: {.callout-note title="Levels"}

In the output above, the `Levels: positive neutral negative` indicates that the returned vector of classifications is a `factor` (not a character) vector.
The levels are thus only showing what the possible values printed in the line above are ☺️

::: 

## More complex response formats

So far, we use a very simple response format: the LLMs response must be one of the allowed labels.
However, in text classification and other application, it can often be helpful to more information than the chosen classification label.

Below, we will create a new response format that combines classification choice with a text the LLM needs to generate before naming its classification decision -- which we call _reasoning_.
To create this reasoning object we

1. define an **response object type** using `ellmer`'s `type_object()` function
2. inside the object type, we define two fields:
   - `reasoning`: a _string_ field where the LLM writes its reasoning
   - `category`: an _enumeration_ field where the LLM chooses one of the allowed categories (as before)


```{r}
response_format <- type_object(
  reasoning = type_string(
    description = "Your reasoning of what sentiment category should be assigned to the text"
  ),
  category = type_enum(
    c("positive", "neutral", "negative"), 
    description = "The sentiment category you assign to the text"
  ),
  .description = "Sentiment classification with reasoning"
)
```

If we now pass a text to be classified, the LLM will respond with an object that contains both its reasoning and the chosen category:

```{r}
#| echo: false
model$set_turns(list())
```

```{r}
# NOTE: we re-use the instructions defined above
model$set_system_prompt(instructions)

text <- "I love programming in R!"

# generate a structured response 
model$chat_structured(text, type = response_format)
```

This time, we get a list back with two elements: `reasoning` and `category`.
The element `reasoning` contains the LLMs reasoning text, and `category` contains the chosen sentiment label.

# Conclusion

This is all you need to know to make your first steps in zero-shot text classification with LLMs!
When applying the code covered in this notebook, remember to 

1. Integrate your Hugging Face API key in the `.Renviron` file (if not yet done)
2. Use the `huggingface_chat` function and the ID of an available LLM (see [here](https://huggingface.co/inference/models)) to create a `Chat` object that allows you "talking" to the chosen LLM
3. Reset the conversation history before starting a new classification task using  `model$set_turns(list())`.
4. Define clear instructions for the classification task and response behavior using either a system message or user message.
5. Define a response format that constrains at least the allowed categories.
6. Use `parallel_chat_structured()` to classify multiple texts applying the same instructions.

