---
title: "Discovering word usage differences"
subtitle: 'VU 402150 "Introduction to Computational Text Analysis with R"'
author: Hauke Licht
institute: University of Innsbruck
date: 2026-04-14
toc: true
toc-depth: 2
toc-location: right
toc-expand: true
number-sections: true
number-depth: 1
execute:
  echo: true
  eval: true
  message: false
  warning: false
  fig-height: 2.8
  fig-width: 4.5
code-annotations: select
format:
  html:
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
---

```{r setup}
#| echo: false
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(scipen = 1, digits = 4) #set to two decimal 
```


::: {.callout-note title="a simple question"}

how do groups of texts differ in their language use

:::

- document-term count matrices provide insights into _lexical_ differences $\Rightarrow$ differences in word usage **frequencies**
- word use difference statistics gives insights into **most "distinctive" terms**

::: {.callout-note title="Monroe et al."}

Monroe et al. ([2017](https://doi.org/10.1093/pan/mpn018))

- discuss different commonly applied statistics 
- propose a new, statistically more sound one

:::



::: {.callout-tip title="Running example"}

Differences between word usage of Labour and Convervative parties in their election manifestos

:::

## Setup

### Load required libraries

```{r libraries}
library(readr)
library(quanteda)
library(quanteda.textmodels)
library(dplyr)
library(ggplot2)
```

```{r}
#| echo: false 
library(ggplot2)
set_theme(
  theme_minimal() + 
    theme(
      theme(
        panel.background = element_rect(fill = "transparent", color = NA_character_),
        # panel.grid.major = element_blank(), # get rid of major grid
        # panel.grid.minor = element_blank(), # get rid of minor grid
        plot.background = element_rect(fill = "transparent", color = NA_character_),
        legend.background = element_rect(fill = "transparent"),
        legend.box.background = element_rect(fill = "transparent"),
        legend.key = element_rect(fill = "transparent")
    )
  )
)
```


### Load and prepare corpus

```{r}
data(
	"data_corpus_ukmanifestos", 
	package = "quanteda.corpora"
)
toks <- tokens(
	data_corpus_ukmanifestos, 
	remove_punct=TRUE, 
	remove_symbols=TRUE,
	remove_numbers=TRUE
)
toks <- tokens_remove(toks, stopwords("en"))
dtm <- dfm(toks, tolower=TRUE)
```


```{r}
# step 1: load corpus and prepare DTM 
data("data_corpus_ukmanifestos", package = "quanteda.corpora")
toks <- tokens(
	data_corpus_ukmanifestos, 
	remove_punct=TRUE, 
	remove_symbols=TRUE,
	remove_numbers=TRUE
)
toks <- tokens_remove(toks, stopwords("en"))
dtm <- dfm(toks, tolower=TRUE)
```

```{r}
# step 2: subset to selected parties' manifestos
dtm <- dfm_subset(x = dtm, subset = docvars(dtm, "Party") %in% c("Lab", "Con")
)
```


let's pick two parties to compare

```{r}
#| eval: false
head(docvars(dtm), 3)
```

```{r}
#| echo: false
head(docvars(dtm)[1:5], 3)
```

```{r}
grp_var <- "Party"
parties <- c("Lab", "Con")
# subset to selected parties' manifestos
dtm <- dfm_subset(
  x = dtm, 
  subset = docvars(dtm, grp_var) %in% parties
)
```

aggregate documents at **_group_ level** (i.e., one row per party)

```{r}
grouped_corp <- dfm_group(
	x = dtm, 
	groups = docvars(dtm, grp_var)
)
ndoc(grouped_corp)
docnames(grouped_corp)
```

```{r}
grouped_corp <- grouped_corp[match(parties, docnames(grouped_corp)), ]  # ensure row order: Labour, Conservative
```

### Load helper functions

```{r}
source(file.path("R", "token_distinctiveness.R"))
# imports custom helper functions
# - get_top_n_terms() 
# - plot_scores()
```

# metrics

## Word count differences

Let's compute "raw" word count differences:

```{r}
# step 1: get tokens counts matrix
tok_counts <- as.matrix(grouped_corp)
tok_counts[, 1:5]
```

```{r}
# step 2: subtract counts between groups
diff_freq <- tok_counts["Lab", ] - tok_counts["Con", ]
diff_freq[1:4]
```

```{r}
stats <- get_top_n_terms(diff_freq, groups = parties)
plot_scores(x = stats, xlab = "Difference in word counts")
```

### Cons

- Does not control for total word counts (makes groups incomparable when one has more text)
- dominated by high-frequency function words (“the,” “and,” “to”) $\Rightarrow$ without additional term weighting or stopword removal, common but uninformative words dominate the top differences
- still ignores _sampling variability_


::: {.callout-note title="sampling variability"}

- word counts (and their differences) are based on a _finite_ sample
- no accounting for fact that resulting metrics are noisy estimates of 'true' underlying word-use probabilities
- *sampling variability* is the _expected_ variation in estimates if multiple samples were drawn (e.g., different documents)

:::

## Word proportion differences

Let's compute _relative_ word use differences:

```{r}
# step 1: compute total tokens per group
toks_per_group <- rowSums(tok_counts)

# step 2: compute proportions per group
tok_props <- tok_counts/toks_per_group

# step 3: subtract proportions between groups
diff_prop <- tok_props["Lab", ] - tok_props["Con", ]
diff_prop[1:4]
```

```{r}
stats <- get_top_n_terms(diff_prop, groups = parties)
plot_scores(x = stats, xlab = "Difference in word proportions")
```


### Pros and cons

Pros: 

- Controls for total word counts
- makes groups comparable even if one has more text

Cons: 

- still dominated by high-frequency function words (“the,” “and,” “to”) $\Rightarrow$ without additional term weighting or stopword removal, common but uninformative words dominate the top differences
- still ignores _sampling variability_

## Log-odds ratio

```{r}
# add small constant to avoid zero proportions
eps <- 1e-12 
# clip proportions to avoid 0 or 1
tok_props_clipped <- pmin(pmax(tok_props, eps), 1 - eps)
# compute odds
odds <- tok_props_clipped / (1 - tok_props_clipped)
odds_ratio <- odds["Lab", ] / odds["Con", ]
log_odds <- log(odds_ratio)
log_odds[1:4]
```

```{r}
stats <- get_top_n_terms(log_odds, groups = parties)
plot_scores(x = stats, xlab = "Log-odds ratio")
```

### Pros and cons

Pros: 

- the log odds ratio introduces symmetry
- expresses association strength $\Rightarrow$ provides a more interpretable measure of association ("how much more likely a word is used by one group?"). 

Cons: 

- unstable for rare words (without smoothing)
- still ignores _sampling variability_

## Smoothed log-odds ratio

adding a small constant to counts (like 0.5) stabilize estimates for very rare words

```{r}
compute_logit <- function(group, e = 0.5) { 
  counts <- tok_counts[group, ]
  total  <- toks_per_group[group]
  log( (counts + e) / (total - counts + e) ) 
}
logit_L <- compute_logit("Lab")
logit_C <- compute_logit("Con")
log_odds_smooth <- logit_L - logit_C
log_odds_smooth[1:4]
```

```{r}
stats <- get_top_n_terms(log_odds_smooth, groups = parties)
plot_scores(x = stats, xlab = "Log-odds ratio (smoothed)")
```

### Pos and cons

Pros: 

- the log odds ratio introduces symmetry
- expresses association strength $\Rightarrow$ provides a more interpretable measure of association ("how much more likely a word is used by one group?"). 
- smoothing introduces Dirichlet priors to shrink extreme values toward zero, avoiding division by zero and extreme values for rare words

Cons:

- still overemphasizes rare or idiosyncratic words, reducing "semantic validity"
- still ignores _sampling variability_

## Monroe et al.'s _Fightin' Words_ $z$-scores 

- builds on smoothed log-odds
- incorporates an informative prior (empirical Bayes) to shrink a token's estimates toward its overall count
- derived variance formula allows computation of *$z$-scores*
- $z>1.96$ can be considered _significantly_ group-specific

## Monroe et al.'s _Fightin' Words_ $z$-scores 

::: {.callout-tip }

I provide an `quanteda`-compatible R implementation in file ["R/token_distinctiveness.R"](https://github.com/haukelicht/cta_intro_course/blob/main/R/token_distinctiveness.R) loaded already above

:::

```{r}
# NOTE: textstat_fighting_words() provided in R/token_distinctiveness.R
fw_stats <- textstat_fighting_words(
  x = grouped_corp, 
  group.var = "Party", 
  .pairs = "permutations"
)
```

`fw_stats` is a list object

```{r}
str(fw_stats, 1)
```

The elements are he comparison results, eg for Lab vs. Con:

```{r}
fw_stats[["Lab-Con"]] |> 
  select(feature, z_score, Party) |> 
  head(6)
```

in comparison between Labour and Conservative party manifestos

- _positive_ $z$-scores $\Rightarrow$ Labour-specific terms
- _negative_ $z$-scores $\Rightarrow$ Conservative-specific terms
- magnitude of $z$-score $\Rightarrow$ strength of "distinctiveness
- $|z| > 1.96$ $\Rightarrow$ statistically significant distinctiveness


```{r}
top_zscores <- fw_stats[["Lab-Con"]] |> 
  # for each party group
  group_by(Party) |>
  # sort from highest to lowest
  arrange(desc(abs(z_score))) |> 
  # get top 20
  slice_head(n = 20) |> 
  ungroup()

plot_scores(top_zscores, value.var = "z_score", group.var = "Party", xlab = "Fightin' words z-score")
```

## Pros and cons

Pros: 

- combines the log odds ratio with its estimated variance
	- takes variance into account
	- provides statistical significance for differences
- produces semantically meaningful, stable results
- most robust and interpretable approach among those copmpared

Cons:

- more complex to compute
- possible overfitting or spurious high scores for very rare words if priors are weak or sample size is small
